---
title: "Introduction to CoverageExperiment"
author: 
  - name: Jacques Serizay
    affiliation:
    - Institut Pasteur, Paris
    email: jacques.serizay@pasteur.fr
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('CoverageExperiment')`"
vignette: >
  %\VignetteIndexEntry{Introduction to CoverageExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL
)
```

## Introduction 

## `CoverageExperiment` class

```{r}
library(CoverageExperiment)
showClass("CoverageExperiment")

data(ce)
ce

rowData(ce)

rowRanges(ce)

colData(ce)

assays(ce)

assay(ce, 'mean')

assay(ce[1, 1], 'mean')
```

### Creating an `CoverageExperiment` object

```{r}
library(rtracklayer)
bw_file <- system.file("extdata", "MNase.bw", package = "CoverageExperiment")
bed_file <- system.file("extdata", "TSSs.bed", package = "CoverageExperiment")
CE <- CoverageExperiment(
    tracks = import(bw_file, as = "Rle"), 
    features = import(bed_file),
    width = 3000, scale = TRUE, center = TRUE
)
CE
```

### Aggregating over multiple tracks / feature sets

```{r}
library(purrr)
library(plyranges)

# ~~~~~~~~~~~~~~~ Import genomic features into a named list ~~~~~~~~~~~~~~~ #
features <- list(
    TSSs = system.file("extdata", "TSSs.bed", package = "CoverageExperiment"),
    `Convergent transcription` = system.file("extdata", "conv_transcription_loci.bed", package = "CoverageExperiment")
) |> map(import) |> map(filter, strand == '+') 

# ~~~~~~~~~~~~~~~ Import coverage tracks into a named list ~~~~~~~~~~~~~~~ #
tracks <- list(
    Scc1 = system.file("extdata", "Scc1.bw", package = "CoverageExperiment"), 
    RNA_fwd = system.file("extdata", "SRR2045244.fwd.bw", package = "CoverageExperiment"),
    RNA_rev = system.file("extdata", "SRR2045244.rev.bw", package = "CoverageExperiment"),
    PolII = system.file("extdata", "PolII.bw", package = "CoverageExperiment"), 
    MNase = system.file("extdata", "MNase.bw", package = "CoverageExperiment")
) |> map(import, as = 'Rle')

# ~~~~~~~~~~~~~~~ Compute aggregated coverage ~~~~~~~~~~~~~~~ #
CE <- CoverageExperiment(tracks, features, width = 5000, scale = TRUE, center = TRUE)
CE
```

### Plotting aggregated coverage with `ggplot2`

```{r}
library(ggplot2)
as_tibble(CE) |> 
    ggplot(aes(x = coord, y = mean)) + 
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high, fill = track), alpha = 0.2) + 
    geom_line(aes(col = track)) + 
    facet_grid(features ~ .) + 
    labs(x = 'Distance from genomic feature', y = 'Mean coverage') + 
    theme_bw() + 
    theme(legend.position = 'top')
```

### Using a tidy grammar

```{r}
library(tidySummarizedExperiment) 
CE |> 
    filter(track != 'Scc1') |> 
    filter(features == 'Convergent transcription') |> 
    ggplot(aes(x = coord, y = mean)) + 
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high, fill = track), alpha = 0.2) + 
    geom_line(aes(col = track)) + 
    facet_grid(track ~ .) + 
    labs(x = 'Distance from locus of convergent transcription', y = 'Mean coverage') + 
    theme_bw() + 
    theme(legend.position = 'top')
```

### Coarsening the aggregated coverage statistics

```{r}
CE2 <- S4Vectors::aggregate(CE, bin = 200)
CE2 |> 
    filter(track != 'Scc1') |> 
    filter(features == 'Convergent transcription') |> 
    ggplot(aes(x = coord, y = mean)) + 
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high, fill = track), alpha = 0.2) + 
    geom_line(aes(col = track)) + 
    geom_point(aes(col = track)) + 
    facet_grid(track ~ .) + 
    labs(x = 'Distance from locus of convergent transcription', y = 'Mean coverage') + 
    theme_bw() + 
    theme(legend.position = 'top')
```

## Example use case: `AnnotationHub` and `TxDb` resources

### Recovering TSSs of forward human genes 

```{r}
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene::TxDb.Hsapiens.UCSC.hg19.knownGene
TSSs <- GenomicFeatures::genes(txdb) |> 
    filter(strand == '+') |> 
    anchor_5p() |> 
    mutate(width = 1)
```

### Recovering H3K4me3 coverage track from ENCODE

```{r}
library(AnnotationHub)
ah <- AnnotationHub()
ah['AH34904']
H3K4me3_bw <- ah[['AH34904']]
```

### Computing the aggregated coverage of H3K4me3 ± 3kb around the TSSs of forward human genes 

```{r}
CE <- CoverageExperiment(
    import(H3K4me3_bw, as = 'Rle'), 
    TSSs, 
    width = 6000, 
    scale = TRUE, center = TRUE
)
ggplot(CE, aes(x = coord, y = mean)) + 
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high, fill = track), alpha = 0.2) + 
    geom_line(aes(col = track)) + 
    facet_grid(track ~ .) + 
    labs(x = 'Distance from TSSs', y = 'Mean coverage') + 
    theme_bw() + 
    theme(legend.position = 'top')
```

### With more genomic tracks

```{r eval = FALSE}
# ~~~~~~~~~~ Recover 15 different histone PTM ChIP-seq tracks ~~~~~~~~~~ #
ids <- c(
    'AH35163', 'AH35165', 'AH35167', 'AH35170', 'AH35173', 'AH35176', 
    'AH35178', 'AH35180', 'AH35182', 'AH35185', 'AH35187', 'AH35189', 
    'AH35191', 'AH35193', 'AH35196'
)
names(ids) <- mcols(ah[ids])$title |> 
    gsub(".*IMR90.", "", x = _) |> 
    gsub("\\..*", "", x = _)
bws <- map(ids, ~ ah[[.x]]) |> 
    map(resource) |> 
    BigWigFileList()
names(bws) <- names(ids)

# ~~~~~~~~~~ Computing coverage over TSSs ~~~~~~~~~~ #
CE <- CoverageExperiment(
    bws, GRangesList(TSSs = TSSs), 
    width = 4000, 
    scale = TRUE, center = TRUE
)

# ~~~~~~~~~~ Plot the resulting CoverageExperiment object ~~~~~~~~~~ #
ggplot(CE, aes(x = coord, y = mean)) + 
    geom_ribbon(aes(ymin = ci_low, ymax = ci_high, fill = track), alpha = 0.2) + 
    geom_line(aes(col = track)) + 
    facet_grid(track ~ .) + 
    labs(x = 'Distance from TSSs', y = 'Mean coverage') + 
    theme_bw() + 
    theme(legend.position = 'top')
```
